---
title: 'EDS 223: assignment 2'
author: "Flora Hamilton"
output: html_document
date: "2023-10-27"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

# [Environmental Justice, Historical Redlining, and Biodiversity Observations]{.underline} 

# A Historical Overview 

Contemporary environmental justice issues often trace their roots back to historical injustices.
The United States, with its enduring legacy of racial segregation, still bears visible traces of this problematic history.

## Historical Context: Redlining and Its Impact

In the 1930s, as part of the New Deal, the Home Owners' Loan Corporation (HOLC) implemented a neighborhood rating system (A (green), B (blue), C (yellow), D (red)) to assess safety for real estate investment.
This system, commonly known as "redlining," had profound consequences by denying access to loans for home ownership[^1].

[^1]: Gee, G. C.
    (2008).
    A multilevel analysis of the relationship between institutional and individual racial discrimination and health status.
    American journal of public health, 98(Supplement_1), S48-S56.

The repercussions of redlining extend beyond economic aspects, impacting community wealth and health.
Redlined neighborhoods exhibit lower greenery[^2] and higher temperatures compared to other areas[^3]
.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021).
    Redlines and greenspace: the relationship between historical redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N.
    (2020).
    The effects of historical housing policies on resident exposure to intra-urban heat: a study of 108 US urban areas.
    Climate, 8(1), 12.

-   Reference:
    -   Gee, G. C.
        (2008).
        A multilevel analysis of the relationship between institutional and individual racial discrimination and health status.
        *American Journal of Public Health, 98*(Supplement_1), S48-S56.

    -   Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
        (2021).
        Redlines and greenspace: the relationship between historical redlining and 2010 greenspace across the United States.
        *Environmental Health Perspectives, 129*(1), 017006.

    -   Hoffman, J. S., Shandas, V., & Pendleton, N.
        (2020).
        The effects of historical housing policies on resident exposure to intra-urban heat: a study of 108 US urban areas.
        *Climate, 8*(1), 12.

    -   [New York Times Coverage](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html)

## Biodiversity and Redlining: An Overlooked Connection

Recent research has uncovered a broader impact of redlining on biodiversity observations.
Community or citizen science, where individuals contribute species observations, generates extensive data.
However, a study by Ellis-Soto and co-authors reveals that redlined neighborhoods remain significantly undersampled across 195 US cities, raising concerns about the implications for conservation decisions[^4].

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H.
    (2023).
    Historical redlining is associated with increasing geographical disparities in bird biodiversity sampling in the United States.
    Nature Human Behaviour, 1-9.

-   Reference:
    -   Ellis-Soto, D., Chapman, M., & Locke, D. H.
        (2023).
        Historical redlining is associated with increasing geographical disparities in bird biodiversity sampling in the United States.
        *Nature Human Behaviour, 1-9*.

    -   [EOS Coverage](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw)

# An Overview of the Workflow 

The goals of this post are to:

\- Investigate the legacy of redlining in current environmental (in)justice \
- Investigate the legacy of redlining in biodiversity observations

The analysis will utilize EJScreen data, acquired from the US EPA, to determine the local health risks communities face in regions of interest -- we designate the geographical context by HOLC grades, and use The [Global Biodiversity Information Facility](gbif.org) to source biodiversity observations.
Observations include a location and date that a species was observed.

**Spatial Skills:**

-   Load vector/raster data

-   Sub-setting

-   Spatial Statistics

-   Transformations / Projections

-   Graphing Results

# More About the Data

## EJScreen

This project will be working with data from the United States Environmental Protection Agency's EJScreen: Environmental Justice Screening and Mapping Tool.

According to the [US EPA website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents or other stakeholders as they search for environmental or demographic information.
> It can also support a wide range of research and policy goals.
> The public has used EJScreen in many different locations and in many different ways.
>
> EPA is sharing EJScreen with the public:\
> - to be more transparent about how we consider environmental justice in our work,\
> - to assist our stakeholders in making informed decisions about pursuing environmental justice and,\
> - to create a common starting point between the agency and the public when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and [block group](https://en.wikipedia.org/wiki/Census_block_group) levels.
I will be working with block group data that has been downloaded from the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data).
To understand the associated data columns, I've read the Technical Documentation and column description spreadsheet available in the data folder and explored the [limitations and caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen) of the data.

#### Mapping Inequality

A team of researchers, led by the [Digital Scholarship Lab](https://dsl.richmond.edu/) at the University of Richmond have digitized maps and information from the HOLC as part of the [Mapping Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58) project.

We will be working with maps of HOLC grade designations for Los Angeles.
Information on the data can be found [here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023, <https://dsl.richmond.edu/panorama/redlining/>

#### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest aggregator of biodiversity observations in the world.
Observations typically include a location and date that a species was observed.

We will be working observations of birds from 2021 onward.

# Part 1: Investigating the legacy of redlining in current environmental (in)justice

Loading relevant packages.

```{r include=TRUE, eval=TRUE}

library(tidyverse)
library(sf)
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
library(plotly)
library(ggplot2)
library(ggspatial)
library(patchwork)
library(tmap)

```

Reading in EJScreen data and filter to Los Angeles County

```{r include=TRUE, eval=TRUE}
library(tidyverse)
library(sf)

# read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- st_read("~/dev/eds223/assignment-2-floraham/Data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb") 

# filter to a state of interest
california <- ejscreen %>%
  filter(ST_ABBREV == "CA") 

# filter to a county of interest 
LA_County <- ejscreen %>%
  filter(CNTY_NAME %in% c("Los Angeles County"))

# find the average values for all variables within counties
california_counties <- aggregate(california, by = list(california$CNTY_NAME), FUN = mean)

# find the centroid within each Census Block Group
# this is helpful to plot Census Block Group data as points
california_blockgroup_centroids <- st_centroid(california)
LA_blockgroup_centroids <- st_centroid(LA_County)
```

To start, I make a map of wastewater discharge by census block groups.
I indicate which census block groups are above the 95th percentile of national values for wastewater discharge by adding a centroid.

```{r }
# name for wastewater discharge pwdis
#wastewater_95 = LA_County |> filter(P_PWDIS > 95)
centroid_pwdis95 <- LA_blockgroup_centroids |>  filter(P_PWDIS > 95)
tm_shape(LA_County) + tm_graticules(labels.size = 0.3) + tm_title("Wastewater Discharge (PWDIS) by Census Blockgroup", size =1) + tm_fill("P_PWDIS", title = "PWDIS (Wastewater Discharge)", size =0.5) + tm_shape(centroid_pwdis95) + tm_bubbles(size = 0.2 , fill = "white", col = "orange") + tm_scalebar(position = c("RIGHT", "BOTTOM")) + tm_compass(position = c("LEFT", "BOTTOM")) + tm_xlab("Longitude", size = 0.5) + tm_ylab("Latitude", rotation = 90, size = 0.5) + tm_add_legend(title = "Census block groups", symbol = "centroid_pwdis95", labels = "Above 95th %ile in PWDIS", col = "orange", shape = 1, col = "white")


```

Finding the percent of census block groups that have:\
- less than 5% of the population is considered low income

```{r }
centroid_li5 <- LA_blockgroup_centroids |>  filter(LOWINCPCT < 0.05)

print(paste(round(nrow(centroid_li5)/nrow(LA_blockgroup_centroids)*100,2), "%"))
```

Finding the percent of census block groups that are:\
- above the 80th percentile for Particulate Matter 2.5 AND\
- above the 80th percentile for Superfund proximity

```{r }

centroid_pm25 <- LA_blockgroup_centroids |>  filter(P_PM25 > 80 & P_PNPL > 80)

print(paste(round(nrow(centroid_pm25)/nrow(LA_blockgroup_centroids)*100,3), "%"))
```

Import redlining information for Los Angeles.

```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/downloads/geojson/CALosAngeles1939.geojson") %>% 
  st_make_valid()

```

I then make a map of historical redlining boundaries, colored by HOLC grade.

```{r }
 tm_shape(LA_redlining)  + tm_graticules() + tm_fill("holc_grade", title = "HOLC Grade", col = "red") + tm_shape(LA_County)  + tm_borders(col = "grey", col_alpha = 0.3) + tm_title( "Historical Redlining Boundaries, Colored by HOLC Grade") + tm_scalebar(position = c("RIGHT", "BOTTOM")) + tm_compass(position = c("LEFT", "BOTTOM")) + tm_xlab("Longitude", size = 0.5) + tm_ylab("Latitude", rotation = 90, size = 0.5) 

```

Next, I will find the number of census block groups that fall within areas with HOLC grades, making sure the CRS match.

```{r eval=TRUE}
## Check CRS of objects: 
st_crs(LA_County) ## WGS 84 PseudoMercator 
#st_crs(LA_blockgroup_centroids) #WGS 84 PsuedoMercator 
st_crs(LA_redlining) #WGS 84 (Does not specify Pseudomercator?)

#transform both CRS's to be the same between blockgroup_centroids and redlining
redlining_scf_wgs = st_transform(LA_redlining, crs = st_crs(LA_County))
  
redlining_sfc_wgs = redlining_scf_wgs  %>% st_set_crs(st_crs(LA_blockgroup_centroids))

##check same CRS
st_crs(redlining_sfc_wgs) == st_crs(LA_blockgroup_centroids)


###  now find intersection 
intersected <- st_intersection(redlining_sfc_wgs, LA_County)

#count the number of rows to check the number of census blockgroups that fit this requirement 
nrow(intersected)

### there are 6090 number of blockgroups that fall within areas with HOLC grades

```

I will summarize current conditions based on EJScreen data within historical redlining categories using the mean of the following variables:\
- % low income.\
- percentile for particulate Matter 2.5.\
- percentile for low life expectancy.\
- percentile for air toxics cancer risk

```{r }
intersected_grouped <- intersected %>% group_by(holc_grade) %>% summarise(mean_low_inc = mean(LOWINCPCT, na.rm = TRUE), 
                                                                          perc_PM25 = mean(P_PM25, na.rm = TRUE),
                                                                          perc_P_LIFEEXPPCT = mean(P_LIFEEXPPCT, na.rm = TRUE),
                                                                          perc_cancer = mean(P_CANCER, na.rm = TRUE))
View(intersected_grouped)

```

### Results Discussion 

By evaluating mean percentile distributions across historical redlining categories of the variables: % low income , PM2.5 , life expectancy percentile, and air toxics cancer risk percentile, we can the see disproportionate effects of these variables across HOLC grade categories.
Through the data, we see the following patterns: the proportion of low income households increase while HOLC grades decline, percentile PM2.5 risk levels increase as HOLC grades decline, percentile for low life expectancy increases as HOLC grades decline, and cancer risk increases while HOLC grades decline.
The results align with my expectations, and seem to show a negative correlation between HOLC grade levels and health risks across all variables: as HOLC grades decline, health risks increase for those living there.

# Part 2: Investigating the legacy of redlining in biodiversity observations

For bird observations from 2022 that fall within neighborhoods with HOLC grads, find the percent of observations within each redlining categories and plot results.
hint: make sure that the bird observations have the same CRS as redlining data.

```{r }
birds <- read_sf('~/dev/eds223/assignment-2-floraham/Data/gbif-birds-LA')

## Check CRS equivalence
st_crs(birds) == st_crs(redlining_sfc_wgs)


## transform bird to be the same CRS
birds = st_transform(birds, crs = st_crs(LA_County))


## Check CRS equivalence, now it's the same!
st_crs(birds) == st_crs(redlining_sfc_wgs)

#For bird observations from 2022 that fall within neighborhoods with HOLC grads, find the percent of observations within each redlining categories and plot results.

birds_2022 <- birds[birds$year == 2022, ]

bird22_intersect <- st_intersection(redlining_sfc_wgs, birds_2022)
View(bird22_intersect)

####
bird22_intersect_sum <- bird22_intersect %>% group_by(holc_grade) %>% summarise(perc_obs_inHOLC = n()/nrow(bird22_intersect)*100)
View(bird22_intersect_sum)

### PLOT PLOT PLOT #### 
x = bird22_intersect_sum$holc_grade
y = bird22_intersect_sum$p_obs_inHOLC

bird22_intersect_sum %>% ggplot(aes(x=holc_grade, y=perc_obs_inHOLC)) + geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Bird Observation Distribution by HOLC grade",
       x = "HOLC Grade",
       y = "% Observations in HOLC Category")
```

### Results discussion 

The results show that a higher proportion of observations occur in lower HOLC grade locations.
This is somewhat counter-intuitive, as it is easy to assume that birds would "prefer" rural and green spaces (which are more prevalent in high income neighborhoods than low income neighborhoods) to highly dense urban areas.
However, urban zones often provide high food availability, more areas of shelter, and lower predation pressure, creating more habitable environments for some bird species -- especially pigeons, sparrows, blackbirds, crows, and swallows.
The observations, though, should not be conflated with species richness.
The abundance of birds in urban areas could increase the health and safety risk of the low-income communities in those areas, as bird droppings and feathers carry disease-causing pathogens and are highly corrosive to paint.
